---
title: "Data Programming"
subtitle: "Homework 1"
author: "Anna Miletova"
date: "`r format(Sys.time(), '%e %B %Y')`"
format: 
  html:
    embed-resources: true
execute: 
  warning: false
  messages: false
  echo: true
  error: true
---

Transform the dataframe **automobile**

```{r}
automobile<-mtcars
library(tidyverse)
```

1.  Move rownames to a column **cars.name**, remove rownames.
2.  Add a column named **l/100km** whose values are based on column mpg, then delete column mpg.
3.  Change the **am** column to a factor with levels **Automatic** and **Manual**.

```{r}
automobile <- rownames_to_column(automobile, var = "cars.name")
rownames(automobile) <- NULL

automobile$`lp100km` <- 235.215 / automobile$mpg
automobile <- subset(automobile, select = -mpg)

automobile$am <- factor(automobile$am, levels = c(0, 1), labels = c("Automatic", "Manual"))
```

After the transformation display the data frame as a tibble then answer these questions independently.

```{r}
as_tibble(automobile) %>% print(n = nrow(automobile))
```

4.  What is the average number of gears?
5.  Return the rows with cars with horsepower between 200 and 400.
6.  Which car is most fuel efficient car among 6 cylinder cars?

```{r}
average_gear <- mean(automobile$gear)
print(average_gear)

hp_cars <- automobile[automobile$hp >= 200 & automobile$hp <= 400, ]
as_tibble(hp_cars)

six_cyl_cars <- subset(automobile, automobile$cyl == 6)
most_fuel_efficient_car <- six_cyl_cars[which.min(six_cyl_cars$lp100km), ]
print(most_fuel_efficient_car$cars.name)
```
