---
title: "Data Programming"
subtitle: "Homework 2"
author: "Anna Miletova"
date: "`r format(Sys.time(), '%e %B %Y')`"
format: 
  html:
  embed-resources: true
execute: 
  warning: false
messages: false
echo: true
error: true
---

```{r}
# libs
library(tibble)
library(lubridate)
library(dplyr)
```
  
### Import the csv file from https://www.kaggle.com/datasets/vishnuvarthanrao/windows-store (login is required). Parse the Date column into a date.

```{r}

msft <- read.csv("msft.csv")
msft$Date <- as.Date(msft$Date, format='%d-%m-%Y')

```

### 1. Display the dataframe as a tibble.

```{r}

as_tibble(msft)

```

### 2. How long has the data been collected?

```{r}

earliest_date <- as.Date(msft[which.min(msft$Date), "Date"])
latest_date <- as.Date(msft[which.max(msft$Date), "Date"])

date_difference_days <- as.numeric(difftime(latest_date, earliest_date, units = "days"))

print(paste("Data has been collected for", date_difference_days,"days"))

```

### 3. Among the highest rated apps return the 10 apps (rows) with the highest amount of reviews. Display only the first 3 columns.

```{r}

highest_rated_apps <- msft[order(-msft$No.of.people.Rated), ]
as_tibble(highest_rated_apps[1:10, 1:3])

```

### 4. Show the number of apps created per month.

```{r}

msft$month <- format(msft$Date, "%m")
apps_by_month <- table(msft$month)
print(apps_by_month)
msft <- msft[, !(names(msft) == "month")]

```

### 5. Sort the rows of the dataframe from the least wordy to most wordy app name. Display the first 15.

```{r}

msft$word_number <- sapply(strsplit(as.character(msft$Name), " "), length)
most_wordy <- msft[order(msft$word_number), ]
as_tibble(most_wordy[1:15,-ncol(msft)])

```

