---
title: "Animals’ life span. Factors that play role in the longevity of animals."
subtitle: "Data Programming course project work"
author:
  - name: Anna Miletova (BF1-EN)
    email: "annamiletova@gmail.com, 89231151@student.upr.si"
    affiliation: 
      - University Primorska FAMNIT, Koper, Slovenia
date: "`r format(Sys.time(), '%e %B %Y')`"
format: 
  html:
    embed-resources: true
execute: 
  warning: false
messages: false
echo: true
error: true
---

```{r}
library(dplyr)
library(tibble)
```

## 1. Introduction

<!--
*Introduction / Presentation and description of the problem*
-->

## 2. Dataframes

<!--
*Presentation of the data (present the data by showing the head of your dataframes, showing the types and explaining what is column is/what unit is in.)*
-->

#### In my work I've used two dataframes:

### 1) Animal Information Dataset (further -- "animals DF")

```{r}
  animals <- read.csv("Animal Dataset.csv", header = TRUE, sep = ",")
  
  animals <- select(animals,
      Animal,
      Lifespan..years.,
      Countries.Found,
      Habitat,
      Weight..kg.,
      Diet,
      Social.Structure) %>%
    group_by(Animal) %>%
    slice(1)

  animals <- ungroup(animals)
```

The data in the dataframe was filtered to get only **unique entries** of the information we are interested in according to the chosen topic. The used columns are: 

```{r}
  print(colnames(animals))
```

This dataframe contains the majority of the necessary data for my research, including *habitat, body mass, diet type,* and *social structure* of the animal species, forming the foundation of the work.

```{r}
  as_tibble(animals)
```

### 2) Animals (further -- "classifications DF")

```{r}
  classifications <- read.csv("animals.csv", header = TRUE, sep = ",")
  
  classifications <- select(classifications, 
    Animal.Name,
    Class,
    Order,
    Family,
    Genus)
```

This dataframe was involved to get more **detailed animal's classification info** such as:

```{r}
  print(colnames(classifications) %>% setdiff("Animal.Name"))
```

<!--
### 3) AnAge: Animal Ageing and Longevity Data

```{r}
anage <- read.csv("anage_data.csv", header = TRUE, sep = "\t")

print(colnames(anage))

anage <- select(anage,
                Common.name,
                Phylum,
                Class,
                Order,
                Family,
                Species,
                Metabolic.rate..W.)

#anage <- filter(anage, complete.cases(anage))
```

-->

## 3. Data aggregation

<!--
*Aggregation of data. Show the result as a table (at least 3). Use only methods, that lead to a solution, like user defined functions, mutate, filter, summarise, inner_join, ggplot. Use the tidyverse approach by chaining the above functions and using pipes %>%.*
-->

#### For the further analysis, it was needed to process some data.

1) *Lifespan* values were **inconsistent**, so I got a new column *Average Lifespan (dbl, in years)* using defined function and `mutate`. The same was done with the *Weight..kg.* column. The entries with unknown average lifespan were filtered with `filter`, since this parameter is the key of the whole research.

```{r}
  process_intervals <- function(col) {
    cleaned_values <- numeric(length(col))
    
    intervals <- grepl("-", col)
    dash_split <- strsplit(col[intervals], "-")
    interval_avg <- sapply(dash_split, function(x) mean(as.double(x)))
    cleaned_values[intervals] <- interval_avg
    
    days_values <- grepl("days", col, ignore.case = TRUE)
    days_numbers <- as.double(gsub("[^0-9]+", "", col[days_values]))
    cleaned_values[days_values] <- days_numbers / 365
    
    months_values <- grepl("months", col, ignore.case = TRUE)
    months_numbers <- as.double(gsub("[^0-9]+", "", col[months_values]))
    cleaned_values[months_values] <- months_numbers / 12
    
    remaining_values <- !(intervals | days_values | months_values)
    cleaned_values[remaining_values] <- as.double(gsub("[^0-9]+", "", col[remaining_values]))
    
    return(cleaned_values)
  }
  
  data <- animals %>%
    mutate(Average.Lifespan = process_intervals(Lifespan..years.)) %>%
    mutate(Weight..kg. = process_intervals(Weight..kg.)) %>%
    filter(!is.na(Average.Lifespan))

  data$Lifespan..years. <- NULL
  
  
  # TODO: to fix decimal values in weight
  print(data[7,])
  print(data[8,])
  print(data[19,])
```

2) Merging dataframes to get animal's classification info:

```{r}
  data <- left_join(data, classifications, by = c("Animal" = "Animal.Name"))
```

The result dataframe I have been workin with for my data analysis:

```{r}
  as_tibble(data)
  #data %>% as_tibble() %>% print(n=nrow(data))
```

## 4. Explotary data analysis

<!--
*Visualisation of the data, at least 3 relevant (and
distinct) ggplots, that are not the same type of plot. Explain each
ggplot)*
-->

## 5. Conclusion

<!--
*Interpretation / Conclusion*
-->

## 6. References

(@) Animal Information Dataset ([kaggle.com](https://www.kaggle.com/datasets/iamsouravbanerjee/animal-information-dataset))
(@) Animals Dataset ([kaggle.com](https://www.kaggle.com/datasets/hessahalkhattabi/animals))
(@) AnAge: Animal Ageing and Longevity Data Dataset. ([kaggle.com](https://www.kaggle.com/datasets/mattop/anage-animal-ageing-and-longevity-data/data?select=anage_data.csv))
(@) Peto's paradox ([widipedia.org](https://en.wikipedia.org/wiki/Peto%27s_paradox))
